What is an Ancestral Recombination Graph?

### 2022-03-08 paper.tex
Discussion of SPRs

[We should discuss the relationship between ARG nodes and SPRs
here. Cite \citep{deng2021distribution} too. Maybe we want to make
this a separate section, as it's a subtle and important point.]

>>>>

\section*{ARGs and SPRs}
\citet{wiuf1999ancestry,wiuf1999recombination} recast
the coalescent with recombination
as a process along the genome rather than backwards in time.
In this formulation we begin at one end of a chromosome by
sampling a tree under the standard coalescent, and then reason
about how this tree is modified by recombination events
as we move along the genome. Although this algorithm is
significantly more complicated to simulate than Hudson's
backwards in time approach, it has led to substantial
insights into the process of recombination itself
and to the sequentially Markov Coalescent (SMC)
approximations~\citep{mcvean2005approximating,marjoram2006fast}
which are central to many modern methods.

Under the coalescent with recombination (or SMC) each local
tree along the genome is transformed into the next by
a ``subtree prune-and-regraft'' (SPR)
% FIXME there should be earlier refs for SPRs, right?
operation~\citep{song2003on,song2006properties}.
The number of SPRs required to transform one leaf-labelled tree
into another is known as the SPR distance [cite]
and is computable in O(??) time.

[Discussion of SPRs in an EARG]

The effects of recombination as we move from left-to-right along
the genome in an unsimplified GARG are easy to interpret.
Consider the leftmost tree in Fig~\ref{fig-simplification}.
In the first tree the parent of B is F, and in the second
tree its parent is E. Since both parents are present as
unary nodes in the tree, we simply change one edge in
order to effect the transition.
A more complex transition occurs when we have a regraft operation
that occurs above the current root (i.e., the current oldest
non-unary node). For example, the transition between the
second and third trees corresponds to node D switching
from the path through I to H, and we then must add branches
corresponding to passing through L and meeting at M.

The equivalence between an ARG and a set of local trees
separated by SPRs is often mentioned [e.g. x y z], but
it is not obvious how to produce an ARG given (say) a
collection of newick trees. If the internal nodes are
labelled, then we can simply input the edges for each
tree in turn, which can easily be ``squashed'' down
to a minimal representation. However, internal nodes
are rarely labelled in newick trees, and therefore
the process of contructing a graph is much harder.
In princple, we can contruct an EARG by reasoning
about the SPRs that must has occured between two
adjacent trees and postulating the recombination
nodes that must have occured. However, this process
is computationally expensive [How expensive? is this
what Espalier does?] Additionally, unless some extra
information is provided the time of the recombination
will not be known. Fundamentally, we must include the
unary nodes corresponding to recombination parents
in the trees if we wish to recover an ARG from them.

[ Cite \citep{deng2021distribution} somewhere here.]
